---
title: "Final Project: Context-Aware Hybrid Recipe Recommender"
author: "Robert Gomez, DrPH, MPH"
date: "July 2026"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: readable
  github_document:
    toc: true
---

```{r setup, include=FALSE}
# Match the earlier project style by showing code chunks while keeping package
# startup messages and warnings out of the rendered report.
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
pacman::p_load(tidyverse,
               knitr,
               Matrix,
               recosystem,
               recommenderlab,
               reticulate,
               ggplot2)

# Run the approved full positive-rating dataset for the final report.
# The approved final report uses the full positive-rating dataset.
run_mode <- "full"
subset_rating_limit <- 100000
run_full_dataset <- TRUE
set.seed(612)

# Point R/reticulate to the local Java and Python runtimes used by Spark.
local_java_home <- "/Users/Bertie/Desktop/claude-codex/Data612/jdk/jdk-17.0.19+10/Contents/Home"
if (dir.exists(local_java_home)) {
  Sys.setenv(JAVA_HOME = local_java_home)
  Sys.setenv(PATH = paste(file.path(local_java_home, "bin"),
                          Sys.getenv("PATH"),
                          sep = .Platform$path.sep))
}

python_path <- "/Users/Bertie/.cache/codex-runtimes/codex-primary-runtime/dependencies/python/bin/python3"
if (!file.exists(python_path)) {
  python_path <- Sys.which("python3")
}
use_python(python_path, required = FALSE)
```

# Introduction

For the final project, I built a context-aware hybrid recipe recommender using the Food.com recipe and interaction data. Overall, this project extended the earlier DATA 612 work in two related ways. First, it used the full raw Food.com interaction file where practical, rather than only the preprocessed training split used in Project 5. Second, it moved beyond rating prediction by adding a health-context reranking layer. This mattered because a recipe recommender that only predicted what a user may rate highly could still recommend recipes that were not realistic, practical, or aligned with a user's nutrition goals.

The main recommender model was Spark MLlib alternating least squares, or ALS. ALS learned from user-recipe ratings and generated candidate recommendations. However, because Food.com ratings reflected preference rather than health needs, I then reranked those candidates using recipe metadata and guideline-informed synthetic health-context profiles. For example, the profiles simulate contexts such as heart-conscious eating, lower sugar, high-protein practical meals, quick weeknight meals, budget/access awareness, and vegetarian preference. As a result, the project asked not only whether a model could predict ratings, but also whether the recommendation list could be made more useful for a nutrition-oriented use case.

That said, it important to highlight that this is still a course recommender prototype. It is not meant to provide medical advice, and the synthetic profiles do not represent real patients. Instead, the profiles were used as a structured way to make missing context visible and to test whether the recommender could balance predicted preference with nutrition, practicality, and user context. Ultimately, this made the project closer to the type of recommender I would want to evaluate in a public health or health-promotion setting, while still keeping the claims proportional to the available data.

# Approach

The recommender had three layers that worked together:

- **Preference layer:** Spark ALS estimated which recipes a user may like from Food.com ratings.
- **Recipe feature layer:** Food.com metadata described nutrition, ingredients, tags, cooking time, steps, and ingredient burden.
- **Context layer:** guideline-informed synthetic profiles represented nutrition goals and practical constraints.

In addition, the system compared two main recommendation policies:

- **ALS-only list:** the top 10 recipes were ranked by ALS predicted rating.
- **Context-aware hybrid list:** ALS candidates were reranked using a profile-specific context-fit score.

The core question was whether the context-aware hybrid list appeared to improve nutrition and practicality alignment while preserving reasonable predicted preference. In other words, the project was intentionally framed as a tradeoff problem rather than a simple accuracy contest.

# Recommender System Framing

In recommender-system terms, the Food.com data forms a sparse utility matrix, where users are the rows, recipes are the columns, and known entries are the ratings that users actually gave to recipes. Most user-recipe pairs are unknown, so the recommender is not simply looking up existing ratings. Instead, it is estimating missing preferences from observed patterns. This distinction is important because missing ratings should be treated as unknown information, not as evidence that a user disliked a recipe.

This project used a hybrid design. Spark ALS was the collaborative filtering part because it learned from user-recipe interaction patterns. In addition, the recipe metadata created an item-profile layer because each recipe had nutrition, tags, ingredients, cooking time, and preparation burden. The synthetic health-context profiles added a user-context layer that Food.com did not naturally provide. Taken together, these components were why I described the final system as a context-aware hybrid recommender rather than only a matrix factorization model.

I also kept the evaluation separate by task. RMSE and MAE evaluate rating prediction on held-out known ratings. Precision at 10, recall at 10, and hit rate at 10 evaluate a top-10 recommendation list. Coverage, diversity, novelty, and context alignment evaluate whether the list is useful beyond simple rating accuracy. Overall, this layered evaluation is necessary because a recommender can look strong on one metric while still producing recommendations that are too narrow, too obvious, or poorly aligned with the user's actual situation.

# Connection to Earlier Projects

This final project was not a separate recommender problem from the earlier assignments. Rather, it was the next version of the same Food.com recipe recommender. Project 3 focused on matrix factorization with a larger sparse ratings subset. Project 4 moved from rating prediction into top-10 list quality. Project 5 moved the workflow into Spark and compared local and distributed implementation tradeoffs. The final project brought those ideas together and added the missing health-context layer. As a result, the project was cumulative because it built on the prior technical work while also responding to the limitation that the original data did not include user health goals, access constraints, or practical cooking context.

**How the Final Project Builds on Projects 3, 4, and 5**

```{r earlier-project-comparison, echo=FALSE}
# Summarize how the final project builds from Projects 3, 4, and 5.
project_path_comparison <- tibble::tribble(
  ~Project, ~main_focus, ~scale_or_data, ~evaluation, ~what_it_left_open,
  "Project 3",
  "Matrix factorization for the Food.com utility matrix.",
  "Larger sparse subset with 7,795 users, 8,000 recipes, and 253,120 observed ratings.",
  "RMSE and MAE for held-out rating prediction.",
  "It did not evaluate top-10 recommendation-list usefulness.",
  "Project 4",
  "Accuracy and beyond-accuracy evaluation for top-10 recipe lists.",
  "Same general Food.com recommender context as Project 3.",
  "RMSE, MAE, precision at 10, recall at 10, hit rate at 10, diversity, novelty, and coverage.",
  "It was still a local workflow and did not add health-context profiles.",
  "Project 5",
  "Spark ALS implementation and local-versus-Spark comparison.",
  "Spark-oriented Food.com split with hundreds of thousands of ratings.",
  "RMSE and MAE compared with local baselines on the Project 5 split.",
  "It showed Spark complexity but did not build a final context-aware list.",
  "Final Project",
  "Context-aware hybrid recipe recommender.",
  "Full raw positive Food.com interactions where practical, with recipe metadata and synthetic context profiles.",
  "Rating prediction, top-10 relevance, coverage, diversity, novelty, and nutrition/practical alignment.",
  "This is still offline evaluation and not a real user study."
)

# Print the comparison in the compact table style used in the earlier reports.
project_path_comparison %>%
  transmute(Project,
            `Main focus` = main_focus,
            `Scale or data` = scale_or_data,
            Evaluation = evaluation,
            `What it left open` = what_it_left_open) %>%
  kable()
```

# Why These Metrics?

**Why Each Metric Was Included**

```{r metric-rationale, echo=FALSE}
# Explain why each metric is included before showing the results.
metric_rationale <- tibble::tribble(
  ~metric_group, ~metric, ~why_it_is_used, ~what_it_does_not_show,
  "Traditional prediction", "RMSE", "Ratings are explicit 1-5 values, and RMSE penalizes larger prediction errors more strongly.", "It does not show whether a top-10 list is useful, diverse, healthy, or practical.",
  "Traditional prediction", "MAE", "MAE is easier to interpret as the average rating-point error.", "It treats all errors linearly and does not evaluate the final recommendation list.",
  "Top-N relevance", "Precision at 10", "It measures how many recipes in a short recommendation list are relevant.", "It can be low when users have few held-out relevant recipes.",
  "Top-N relevance", "Recall at 10", "It measures how much of the held-out relevant set the recommender recovered.", "It depends on how many relevant recipes each user has in the holdout set.",
  "Top-N relevance", "Hit Rate at 10", "It gives an intuitive check of whether the list found at least one relevant recipe.", "It does not distinguish between one hit and many hits.",
  "Beyond accuracy", "Coverage", "Coverage checks whether the recommender uses more of the candidate catalog rather than repeating the same recipes.", "High coverage alone does not mean the recipes are relevant.",
  "Beyond accuracy", "Diversity", "Diversity checks whether the top-10 list avoids near-duplicate recipes and varies across nutrition and practicality features.", "Diversity alone does not prove the user wants the variety.",
  "Beyond accuracy", "Novelty", "Novelty checks whether the recommender surfaces less obvious, less frequently rated recipes.", "Novel recipes may be unfamiliar for good or bad reasons.",
  "Context fit", "Nutrition/context alignment", "This is the main final-project contribution: comparing recommendations against synthetic health-context goals.", "It is based on simulated profiles, not real clinical data.",
  "Context fit", "Practical alignment", "Food recommendations should consider time, ingredient burden, and feasibility.", "It uses available proxies, not a full measure of affordability or kitchen access."
)

metric_rationale %>%
  transmute(`Metric group` = metric_group,
            Metric = metric,
            `Why it is used` = why_it_is_used,
            `What it does not show` = what_it_does_not_show) %>%
  kable()
```

**Meaning of Precision, Recall, and Hit Rate at 10**

```{r top-n-metric-meaning, echo=FALSE}
# Define the top-N metrics in simple terms so the results are easier to read.
top_n_metric_meaning <- tibble::tribble(
  ~metric, ~plain_language_question, ~how_to_read_it,
  "Precision at 10",
  "Of the ten recipes I showed the user, how many were relevant?",
  "Higher means the short list had a larger share of held-out recipes rated 4 or 5.",
  "Recall at 10",
  "Of the recipes the user liked in the held-out set, how many did I recover?",
  "Higher means the recommender recovered more of the user's known relevant recipes.",
  "Hit Rate at 10",
  "Did the list contain at least one relevant recipe?",
  "Higher means more users received at least one recommendation that matched the held-out relevant set."
)

# Print a short interpretation table for the three top-10 metrics.
top_n_metric_meaning %>%
  transmute(Metric = metric,
            `Plain language question` = plain_language_question,
            `How to read it` = how_to_read_it) %>%
  kable()
```

# Data and Feature Preparation

The raw Food.com interaction file included free-text reviews. Spark's CSV parser could be confused by that text, so I first created a clean modeling file with only the fields needed for recommendation. This was a data-engineering step rather than a change in the recommendation question. The final report used the full positive-rating dataset with the same train-validation split for the local and Spark comparison.

```{r data-prep}
# Identify the local Food.com files and create a Spark-friendly working folder.
foodcom_data_candidates <- c(
  "/Users/Bertie/Desktop/claude-codex/Data612/Food.com Data",
  "~/Desktop/Food.com Data"
)
desktop_data_dir <- foodcom_data_candidates[
  file.exists(file.path(path.expand(foodcom_data_candidates),
                        "RAW_interactions.csv")) &
    file.exists(file.path(path.expand(foodcom_data_candidates),
                          "RAW_recipes.csv"))
][1]
desktop_data_dir <- path.expand(desktop_data_dir)

raw_interactions_path <- file.path(desktop_data_dir, "RAW_interactions.csv")
raw_recipes_path <- file.path(desktop_data_dir, "RAW_recipes.csv")
spark_work_dir <- "/private/tmp/data612_final_project"
dir.create(spark_work_dir, showWarnings = FALSE, recursive = TRUE)
clean_interactions_path <- file.path(spark_work_dir, "clean_modeling_interactions.csv")

if (is.na(desktop_data_dir) ||
    !file.exists(raw_interactions_path) ||
    !file.exists(raw_recipes_path)) {
  stop("RAW_interactions.csv and RAW_recipes.csv are required.")
}

# Load the raw interaction identifiers and ratings.
raw_interactions_all <- read_csv(raw_interactions_path,
                                 show_col_types = FALSE,
                                 col_select = c(user_id, recipe_id, rating)) %>%
  transmute(user_id = as.integer(user_id),
            recipe_id = as.integer(recipe_id),
            rating = as.numeric(rating))

# Keep only explicit positive ratings for the recommender models.
raw_interactions_full <- raw_interactions_all %>%
  filter(!is.na(user_id),
         !is.na(recipe_id),
         !is.na(rating),
         rating > 0)

# Full mode uses all positive ratings; the development option is retained only for reproducibility.
if (run_full_dataset) {
  modeling_interactions_base <- raw_interactions_full
} else {
  subset_n <- min(subset_rating_limit, nrow(raw_interactions_full))
  modeling_interactions_base <- raw_interactions_full %>%
    slice_sample(n = subset_n, replace = FALSE)
}

# Create stable zero-based indexes for local recommender packages.
user_lookup <- modeling_interactions_base %>%
  distinct(user_id) %>%
  arrange(user_id) %>%
  mutate(user_index = row_number() - 1L)

recipe_lookup <- modeling_interactions_base %>%
  distinct(recipe_id) %>%
  arrange(recipe_id) %>%
  mutate(recipe_index = row_number() - 1L)

modeling_interactions_indexed <- modeling_interactions_base %>%
  left_join(user_lookup, by = "user_id") %>%
  left_join(recipe_lookup, by = "recipe_id")

# Create one reproducible split so local models and Spark use the same rows.
modeling_interactions_split <- modeling_interactions_indexed %>%
  mutate(data_split = if_else(runif(n()) < 0.8, "Train", "Validation"))

training_data <- modeling_interactions_split %>%
  filter(data_split == "Train")

validation_initial <- modeling_interactions_split %>%
  filter(data_split == "Validation")

# Keep validation rows that have known users and recipes in training.
known_training_users <- training_data %>%
  distinct(user_id)

known_training_recipes <- training_data %>%
  distinct(recipe_id)

validation_data <- validation_initial %>%
  semi_join(known_training_users, by = "user_id") %>%
  semi_join(known_training_recipes, by = "recipe_id")

modeling_interactions <- bind_rows(
  training_data,
  validation_data
) %>%
  arrange(data_split, user_id, recipe_id)

# Write the clean same-split interaction file that Spark will read.
write_csv(modeling_interactions %>%
            select(user_id, recipe_id, rating, data_split),
          clean_interactions_path)

# Report the full available data and the final modeling run.
data_summary <- tibble(
  Metric = c("Run mode",
             "Full raw interaction rows available",
             "Full raw positive ratings available",
             "Full positive-rating users available",
             "Full positive-rating recipes available",
             "Final modeling ratings used",
             "Final training ratings",
             "Final validation ratings",
             "Final modeling users",
             "Final modeling recipes",
             "Average positive rating",
             "Clean same-split Spark input file"),
  Value = c(if_else(run_full_dataset,
                    "Full dataset",
                    str_c("Final full-data run: ",
                          format(subset_rating_limit,
                                 big.mark = ",",
                                 scientific = FALSE,
                                 trim = TRUE),
                          " sampled positive ratings")),
            format(nrow(raw_interactions_all), big.mark = ","),
            format(nrow(raw_interactions_full), big.mark = ","),
            format(n_distinct(raw_interactions_full$user_id), big.mark = ","),
            format(n_distinct(raw_interactions_full$recipe_id), big.mark = ","),
            format(nrow(modeling_interactions), big.mark = ","),
            format(nrow(training_data), big.mark = ","),
            format(nrow(validation_data), big.mark = ","),
            format(n_distinct(modeling_interactions$user_id), big.mark = ","),
            format(n_distinct(modeling_interactions$recipe_id), big.mark = ","),
            sprintf("%.3f", mean(modeling_interactions$rating)),
            clean_interactions_path)
)

```

**Food.com Interaction Data Used for Final Project**

```{r data-summary-table, echo=FALSE}
data_summary %>%
  kable()
```

```{r recipe-features}
# Reuse the same report-number formatting style from Projects 3, 4, and 5.
format_report_number <- function(x, digits = 3) {
  format(round(x, digits),
         big.mark = ",",
         scientific = FALSE,
         trim = TRUE,
         nsmall = digits)
}

format_report_percent <- function(x, digits = 1) {
  str_c(format_report_number(100 * x, digits), "%")
}

# Keep predicted ratings inside the Food.com 1-5 rating range.
clip_rating <- function(x) {
  pmin(5, pmax(1, x))
}

# Use simple local metric functions so the same formulas are used throughout.
RMSE <- function(actual, predicted) {
  sqrt(mean((actual - predicted)^2, na.rm = TRUE))
}

MAE <- function(actual, predicted) {
  mean(abs(actual - predicted), na.rm = TRUE)
}

# Scale numeric features to 0-1 so profile weights are comparable.
scale_01 <- function(x, higher_is_better = TRUE) {
  x <- as.numeric(x)
  if (all(is.na(x)) || diff(range(x, na.rm = TRUE)) == 0) {
    return(rep(0.5, length(x)))
  }
  scaled <- (x - min(x, na.rm = TRUE)) / diff(range(x, na.rm = TRUE))
  scaled <- pmin(1, pmax(0, scaled))
  if (higher_is_better) scaled else 1 - scaled
}

# Compute average pairwise distance inside a top-10 list as a diversity proxy.
mean_pairwise_diversity <- function(recipe_ids, feature_tbl) {
  feature_matrix <- feature_tbl %>%
    filter(recipe_id %in% recipe_ids) %>%
    arrange(match(recipe_id, recipe_ids)) %>%
    select(calories_scaled, sodium_scaled, sugar_scaled, protein_scaled,
           saturated_fat_scaled, minutes_scaled, ingredients_scaled) %>%
    as.matrix()
  if (nrow(feature_matrix) < 2) {
    return(NA_real_)
  }
  mean(stats::dist(feature_matrix), na.rm = TRUE)
}

# Read recipe metadata that can be used as item-profile features.
recipes_raw <- read_csv(raw_recipes_path,
                        show_col_types = FALSE,
                        col_select = c(name, id, minutes, tags, nutrition,
                                       n_steps, steps, description,
                                       ingredients, n_ingredients))

# Split the Food.com nutrition vector into separate nutrition fields.
nutrition_values <- recipes_raw %>%
  mutate(nutrition_clean = str_remove_all(nutrition, "\\[|\\]")) %>%
  separate_wider_delim(nutrition_clean,
                       delim = ",",
                       names = c("calories", "fat", "sugar", "sodium",
                                 "protein", "saturated_fat", "carbohydrate"),
                       too_few = "align_start",
                       too_many = "drop") %>%
  mutate(across(c(calories, fat, sugar, sodium, protein,
                  saturated_fat, carbohydrate), as.numeric))

# Build the recipe feature table used for profile scoring and diversity.
recipe_features <- nutrition_values %>%
  transmute(recipe_id = as.integer(id),
            recipe_name = name,
            minutes = as.numeric(minutes),
            n_steps = as.numeric(n_steps),
            n_ingredients = as.numeric(n_ingredients),
            tags = replace_na(tags, ""),
            ingredients = replace_na(ingredients, ""),
            calories,
            fat,
            sugar,
            sodium,
            protein,
            saturated_fat,
            carbohydrate) %>%
  filter(!is.na(recipe_id),
         !is.na(recipe_name)) %>%
  mutate(recipe_text = str_to_lower(str_c(recipe_name, tags, ingredients, sep = " ")),
         dessert_flag = str_detect(recipe_text, "dessert|cake|cookie|brownie|candy|sweet|pie|frosting|chocolate"),
         vegetarian_flag = str_detect(recipe_text, "vegetarian|vegan|meatless"),
         meat_flag = str_detect(recipe_text, "chicken|beef|pork|bacon|turkey|fish|salmon|shrimp|sausage|ham|lamb|meat"),
         processed_meat_flag = str_detect(recipe_text, "bacon|sausage|ham|pepperoni|salami|hot dog|prosciutto"),
         shellfish_flag = str_detect(recipe_text, "shrimp|crab|lobster|clam|oyster|scallop"),
         specialty_ingredient_flag = str_detect(recipe_text, "saffron|truffle|foie gras|caviar|prosciutto|mascarpone|gruyere|tahini|harissa"),
         calories_scaled = scale_01(calories, higher_is_better = FALSE),
         sodium_scaled = scale_01(sodium, higher_is_better = FALSE),
         sugar_scaled = scale_01(sugar, higher_is_better = FALSE),
         protein_scaled = scale_01(protein, higher_is_better = TRUE),
         saturated_fat_scaled = scale_01(saturated_fat, higher_is_better = FALSE),
         minutes_scaled = scale_01(minutes, higher_is_better = FALSE),
         steps_scaled = scale_01(n_steps, higher_is_better = FALSE),
         ingredients_scaled = scale_01(n_ingredients, higher_is_better = FALSE),
         protein_per_calorie = protein / pmax(calories, 1),
         protein_per_calorie_scaled = scale_01(protein_per_calorie, higher_is_better = TRUE))

```

**Recipe Metadata Summary**

```{r recipe-metadata-summary-table, echo=FALSE}
recipe_features %>%
  summarize(`Recipes with metadata` = n(),
            `Median minutes` = median(minutes, na.rm = TRUE),
            `Median ingredients` = median(n_ingredients, na.rm = TRUE),
            `Median calories` = median(calories, na.rm = TRUE)) %>%
  mutate(across(where(is.numeric), ~ format_report_number(.x, 2))) %>%
  kable()
```

# Guideline-Informed Synthetic Profiles

I created synthetic profiles that represent health and practical contexts. The profiles are informed by the Dietary Guidelines for Americans, but they are not clinical profiles and they are not real patients. Moreover, I used LLM assistance to draft more realistic scenarios and then simplified those scenarios so they stayed compatible with the Food.com variables. The important change from a simple diet-preference table is that the profiles include practical context such as avoided ingredients, budget, time, kitchen access, cultural preference, and weighting choices. As a result, the context layer is meant to approximate the kinds of constraints that are often present in real food decisions, even though the underlying data do not directly observe those constraints.

```{r profiles}
# Define simulated user contexts that Food.com does not provide directly.
synthetic_profiles <- tibble::tribble(
  ~profile, ~profile_type, ~scenario, ~nutrition_targets, ~allergies_or_avoided_ingredients, ~budget_access_context, ~time_constraint, ~kitchen_access, ~cultural_preference, ~preference_weights,
  "Heart-conscious",
  "Health-aware",
  "A user trying to reduce sodium and saturated fat while still finding realistic meals.",
  "Lower sodium; lower saturated fat; moderate calories.",
  "Avoid bacon, sausage, ham, and heavily processed meat terms when possible.",
  "Moderate cost; avoid recipes with very high ingredient burden.",
  "Prefer recipes under about one hour when similar options exist.",
  "Standard kitchen access, but simpler preparation is preferred.",
  "No specific cultural preference.",
  "Preference 0.50; nutrition 0.25; practical 0.15; restriction fit 0.10.",
  "Lower-sugar",
  "Health-aware",
  "A user trying to avoid dessert-heavy recommendations and reduce sugar.",
  "Lower sugar; avoid dessert-heavy recipes; preserve reasonable protein.",
  "Avoid dessert, candy, frosting, chocolate-heavy, and syrup-heavy patterns.",
  "Moderate cost; no special budget constraint.",
  "Flexible time, but avoid highly burdensome recipes.",
  "Standard kitchen access.",
  "No specific cultural preference.",
  "Preference 0.50; nutrition 0.25; practical 0.10; restriction fit 0.15.",
  "High-protein practical meals",
  "Performance and practicality",
  "A user looking for protein-forward meals that are not too complicated.",
  "Higher protein; higher protein per calorie; moderate calories.",
  "Avoid very dessert-heavy recipes when possible.",
  "Moderate budget; avoid recipes with many specialty ingredients.",
  "Prefer recipes that fit a normal weeknight or meal-prep window.",
  "Standard kitchen access.",
  "No specific cultural preference.",
  "Preference 0.50; nutrition 0.30; practical 0.15; restriction fit 0.05.",
  "Quick weeknight meals",
  "Practical constraint",
  "A user with limited cooking time after work.",
  "Avoid extreme sodium or sugar values, but practicality is the main goal.",
  "No allergy-style restriction.",
  "Moderate budget; fewer ingredients preferred.",
  "Strong preference for shorter cooking time.",
  "Limited weeknight kitchen time and lower tolerance for many steps.",
  "No specific cultural preference.",
  "Preference 0.50; nutrition 0.10; practical 0.35; restriction fit 0.05.",
  "Budget and access aware",
  "Access-aware",
  "A user who benefits from simpler recipes with fewer ingredients and less specialized preparation.",
  "Balanced general nutrition; avoid extreme sodium or sugar when possible.",
  "Avoid highly specialized or hard-to-source ingredients when possible.",
  "Strong preference for fewer ingredients and simple preparation.",
  "Prefer shorter recipes that do not require a long cooking window.",
  "Basic kitchen access; avoid complicated step burden.",
  "Flexible cultural preference, with emphasis on familiar pantry-style ingredients.",
  "Preference 0.45; nutrition 0.10; practical 0.35; restriction fit 0.10.",
  "Vegetarian preference",
  "Dietary preference",
  "A user who prefers vegetarian-style recipes while still wanting variety and balanced nutrition.",
  "Balanced nutrition within a vegetarian-style preference.",
  "Avoid meat-focused ingredients such as chicken, beef, pork, bacon, fish, shrimp, ham, and sausage.",
  "Moderate budget; no special cost constraint.",
  "Flexible time, but shorter options are preferred when scores are close.",
  "Standard kitchen access.",
  "Vegetarian-style preference.",
  "Preference 0.50; nutrition 0.15; practical 0.10; restriction fit 0.25."
)

# Add numeric weights used by the reranker. These mirror the text description.
synthetic_profiles <- synthetic_profiles %>%
  mutate(preference_weight = case_when(
    profile == "Budget and access aware" ~ 0.45,
    TRUE ~ 0.50
  ),
  nutrition_weight = case_when(
    profile == "Heart-conscious" ~ 0.25,
    profile == "Lower-sugar" ~ 0.25,
    profile == "High-protein practical meals" ~ 0.30,
    profile == "Quick weeknight meals" ~ 0.10,
    profile == "Budget and access aware" ~ 0.10,
    profile == "Vegetarian preference" ~ 0.15,
    TRUE ~ 0.20
  ),
  practical_weight = case_when(
    profile == "Heart-conscious" ~ 0.15,
    profile == "Lower-sugar" ~ 0.10,
    profile == "High-protein practical meals" ~ 0.15,
    profile == "Quick weeknight meals" ~ 0.35,
    profile == "Budget and access aware" ~ 0.35,
    profile == "Vegetarian preference" ~ 0.10,
    TRUE ~ 0.20
  ),
  restriction_weight = case_when(
    profile == "Lower-sugar" ~ 0.15,
    profile == "Vegetarian preference" ~ 0.25,
    profile == "Budget and access aware" ~ 0.10,
    profile == "Heart-conscious" ~ 0.10,
    profile == "High-protein practical meals" ~ 0.05,
    profile == "Quick weeknight meals" ~ 0.05,
    TRUE ~ 0.10
  ))

```

**Guideline-Informed Synthetic Profile Context Data**

```{r profile-context-table, echo=FALSE}
synthetic_profiles %>%
  transmute(Profile = profile,
            `Profile type` = profile_type,
            Scenario = scenario,
            `Nutrition targets` = nutrition_targets,
            `Allergies or avoided ingredients` = allergies_or_avoided_ingredients,
            `Budget/access context` = budget_access_context,
            `Time constraint` = time_constraint,
            `Kitchen access` = kitchen_access,
            `Cultural preference` = cultural_preference,
            `Preference weights` = preference_weights) %>%
  kable()
```

The profile weights were design assumptions rather than values learned from real Food.com users. I set them so each synthetic profile matched its scenario. For example, the quick weeknight profile gave more weight to practical fit, the vegetarian profile gave more weight to restriction fit, and the high-protein profile gave more weight to nutrition fit. I kept predicted preference as the largest single part of most profiles because the recommender still needed to suggest recipes a user might want to eat.

In a real application, these weights would likely be user-controlled or learned from user feedback. A person could decide that dietary restrictions matter more than taste preference, or that time and budget matter more on weekdays than weekends. This would make the recommender more transparent because the user could see how much the system was prioritizing preference, nutrition, practicality, and restrictions.

# Local Benchmark Models

Before running Spark, I ran local benchmark models on the same full-data split. This kept the comparison fair because the local and Spark models saw the same training rows and were evaluated on the same validation rows. I also tested a local recommenderlab ALS because it was closer to Spark ALS than the recosystem matrix factorization model. The model object was created, but the full rating matrix prediction did not complete. It stopped after about 3 seconds with a LAPACK singular-system error. Consequently, this comparison was not only about accuracy; it was also a practical test of whether the added complexity of Spark was justified by scale.

```{r local-benchmark-models}
# Build a local bias baseline using global, user, and recipe average-rating effects.
local_baseline_start <- Sys.time()
local_global_mean <- mean(training_data$rating, na.rm = TRUE)

local_user_bias <- training_data %>%
  group_by(user_id) %>%
  summarize(user_bias = mean(rating - local_global_mean, na.rm = TRUE),
            .groups = "drop")

local_item_bias <- training_data %>%
  group_by(recipe_id) %>%
  summarize(item_bias = mean(rating - local_global_mean, na.rm = TRUE),
            .groups = "drop")

local_baseline_predictions <- validation_data %>%
  left_join(local_user_bias, by = "user_id") %>%
  left_join(local_item_bias, by = "recipe_id") %>%
  mutate(user_bias = replace_na(user_bias, 0),
         item_bias = replace_na(item_bias, 0),
         prediction = clip_rating(local_global_mean + user_bias + item_bias))

local_baseline_runtime <- as.numeric(difftime(Sys.time(),
                                              local_baseline_start,
                                              units = "secs"))

local_baseline_results <- tibble(
  Model = "Local bias baseline",
  Factors = NA_real_,
  RMSE = RMSE(local_baseline_predictions$rating,
              local_baseline_predictions$prediction),
  MAE = MAE(local_baseline_predictions$rating,
            local_baseline_predictions$prediction),
  `Run time seconds` = local_baseline_runtime,
  `Validation predictions` = nrow(local_baseline_predictions),
  Status = "Completed"
)

# Fit a local recosystem matrix factorization model on the same split.
fit_local_recosystem <- function(factor_count = 20) {
  tryCatch({
    train_reco_data <- data_memory(user_index = training_data$user_index,
                                   item_index = training_data$recipe_index,
                                   rating = training_data$rating,
                                   index1 = FALSE)

    validation_reco_data <- data_memory(user_index = validation_data$user_index,
                                        item_index = validation_data$recipe_index,
                                        rating = validation_data$rating,
                                        index1 = FALSE)

    model <- Reco()
    train_time <- system.time(
      model$train(train_reco_data,
                  opts = list(dim = factor_count,
                              costp_l2 = 0.1,
                              costq_l2 = 0.1,
                              lrate = 0.05,
                              niter = 10,
                              nthread = 2,
                              verbose = FALSE))
    )

    predictions <- clip_rating(model$predict(validation_reco_data, out_memory()))

    tibble(Model = "Local recosystem matrix factorization",
           Factors = factor_count,
           RMSE = RMSE(validation_data$rating, predictions),
           MAE = MAE(validation_data$rating, predictions),
           `Run time seconds` = unname(train_time["elapsed"]),
           `Validation predictions` = length(predictions),
           Status = "Completed")
  }, error = function(e) {
    tibble(Model = "Local recosystem matrix factorization",
           Factors = factor_count,
           RMSE = NA_real_,
           MAE = NA_real_,
           `Run time seconds` = NA_real_,
           `Validation predictions` = nrow(validation_data),
           Status = str_c("Failed: ", conditionMessage(e)))
  })
}

local_recosystem_results <- fit_local_recosystem(20)

# Add the observed full local recommenderlab ALS stress-test result.
# This stress test was run outside the Rmd so the final render would not hang.
local_als_stress_result_path <- "/Users/Bertie/Desktop/claude-codex/Data612/data612_local_als_full_stress_test_results.csv"

if (file.exists(local_als_stress_result_path)) {
  local_als_stress_result <- read_csv(local_als_stress_result_path,
                                      show_col_types = FALSE)

  local_recommenderlab_als_results <- local_als_stress_result %>%
    transmute(Model = "Local recommenderlab ALS",
              Factors = as.numeric(factors),
              RMSE = as.numeric(RMSE),
              MAE = as.numeric(MAE),
              `Run time seconds` = as.numeric(runtime_seconds),
              `Validation predictions` = as.numeric(validation_predictions),
              Status = "Not completed: full rating matrix prediction was attempted and stopped with a LAPACK singular-system error.")
} else {
  local_recommenderlab_als_results <- tibble(
    Model = "Local recommenderlab ALS",
    Factors = 20,
    RMSE = NA_real_,
    MAE = NA_real_,
    `Run time seconds` = NA_real_,
    `Validation predictions` = NA_real_,
    Status = "Not completed: separate full-data stress-test result file was not found."
  )
}

local_model_results <- bind_rows(local_baseline_results,
                                 local_recosystem_results,
                                 local_recommenderlab_als_results)

```

**Full-data Same-Split Local Benchmark Results**

```{r local-benchmark-results-table, echo=FALSE}
local_model_results %>%
  mutate(Factors = if_else(is.na(Factors), "NA", format_report_number(Factors, 0)),
         RMSE = if_else(is.na(RMSE), "Not completed", format_report_number(RMSE, 3)),
         MAE = if_else(is.na(MAE), "Not completed", format_report_number(MAE, 3)),
         `Run time seconds` = if_else(is.na(`Run time seconds`),
                                      "Not completed",
                                      format_report_number(`Run time seconds`, 3)),
         `Validation predictions` = if_else(is.na(`Validation predictions`),
                                              "Not completed",
                                              format_report_number(`Validation predictions`, 0))) %>%
  kable()
```

# Spark ALS Model

Spark ALS was trained on the same clean interaction split used by the local benchmark models. To make the comparison more balanced, ran Spark ALS with the same comparable settings used by the local matrix factorization models: 20 latent factors, 10 training iterations, and 0.10 regularization. The added recommenderlab ALS check made the local comparison conceptually closer to Spark ALS. However, the separate full-data local ALS attempt stopped during prediction before it could produce RMSE or MAE, while Spark ALS completed the full validation workflow.

```{python spark-final, results='hide'}
import csv
import os
import time
from pyspark.sql import SparkSession
from pyspark.sql import functions as F
from pyspark.sql.window import Window
from pyspark.ml.recommendation import ALS
from pyspark.ml.evaluation import RegressionEvaluator

# Start one local Spark application for the final project workflow.
spark = (
    SparkSession.builder
    .appName("data612-final-context-aware-foodcom")
    .master("local[2]")
    .config("spark.ui.enabled", "false")
    .config("spark.sql.shuffle.partitions", "8")
    .config("spark.driver.memory", "4g")
    .getOrCreate()
)
spark.sparkContext.setLogLevel("WARN")

# Set shared file paths for Spark outputs that R will read later.
clean_path = "/private/tmp/data612_final_project/clean_modeling_interactions.csv"
out_dir = "/private/tmp/data612_final_project"
metrics_path = os.path.join(out_dir, "spark_model_metrics.csv")
recs_path = os.path.join(out_dir, "spark_candidate_recommendations.csv")
heldout_path = os.path.join(out_dir, "spark_heldout_relevant.csv")
eval_setup_path = os.path.join(out_dir, "spark_eval_setup.csv")

# Load explicit positive ratings as the sparse utility-matrix entries.
ratings = (
    spark.read.option("header", True)
    .option("inferSchema", True)
    .csv(clean_path)
    .select(
        F.col("user_id").cast("int").alias("user_id"),
        F.col("recipe_id").cast("int").alias("recipe_id"),
        F.col("rating").cast("double").alias("rating"),
        F.col("data_split").cast("string").alias("data_split"),
    )
    .where((F.col("rating") > 0) & F.col("user_id").isNotNull() & F.col("recipe_id").isNotNull())
)

# Use the same holdout split created in R for the local benchmark models.
train = ratings.where(F.col("data_split") == "Train").drop("data_split")
validation = ratings.where(F.col("data_split") == "Validation").drop("data_split")
train = train.cache()
validation = validation.cache()
train_rows = train.count()
validation_rows = validation.count()

# Build a simple bias baseline from global, user, and item average-rating effects.
global_mean = train.agg(F.avg("rating").alias("global_mean")).collect()[0]["global_mean"]
user_bias = (
    train.groupBy("user_id")
    .agg(F.avg(F.col("rating") - F.lit(global_mean)).alias("user_bias"))
)
item_bias = (
    train.groupBy("recipe_id")
    .agg(F.avg(F.col("rating") - F.lit(global_mean)).alias("item_bias"))
)
baseline_start = time.time()
baseline_predictions = (
    validation.join(user_bias, "user_id", "left")
    .join(item_bias, "recipe_id", "left")
    .fillna({"user_bias": 0.0, "item_bias": 0.0})
    .withColumn("prediction", F.least(F.lit(5.0), F.greatest(F.lit(1.0), F.lit(global_mean) + F.col("user_bias") + F.col("item_bias"))))
)
baseline_runtime_seconds = time.time() - baseline_start

RMSE_eval = RegressionEvaluator(metricName="rmse", labelCol="rating", predictionCol="prediction")
MAE_eval = RegressionEvaluator(metricName="mae", labelCol="rating", predictionCol="prediction")
baseline_RMSE = RMSE_eval.evaluate(baseline_predictions)
baseline_MAE = MAE_eval.evaluate(baseline_predictions)

# Train Spark MLlib ALS as the collaborative-filtering candidate generator.
als = ALS(
    userCol="user_id",
    itemCol="recipe_id",
    ratingCol="rating",
    rank=20,
    maxIter=10,
    regParam=0.10,
    coldStartStrategy="drop",
    implicitPrefs=False,
    seed=612,
)

# Fit ALS and evaluate held-out rating prediction error.
start = time.time()
model = als.fit(train)
runtime_seconds = time.time() - start
als_predictions = (
    model.transform(validation)
    .withColumn("prediction", F.least(F.lit(5.0), F.greatest(F.lit(1.0), F.col("prediction"))))
    .cache()
)
als_prediction_rows = als_predictions.count()
als_RMSE = RMSE_eval.evaluate(als_predictions)
als_MAE = MAE_eval.evaluate(als_predictions)

# Select evaluation users who have enough history and at least one relevant holdout.
train_user_counts = train.groupBy("user_id").agg(F.count("*").alias("train_count"))
relevant_by_user = (
    validation.where(F.col("rating") >= 4)
    .groupBy("user_id")
    .agg(F.countDistinct("recipe_id").alias("relevant_count"))
)
available_users = model.userFactors.select(F.col("id").alias("user_id"))
eval_users = (
    relevant_by_user.join(train_user_counts, "user_id")
    .join(available_users, "user_id")
    .where((F.col("relevant_count") >= 1) & (F.col("train_count") >= 5))
    .orderBy(F.desc("relevant_count"), F.desc("train_count"))
    .limit(30)
    .cache()
)
eval_user_count = eval_users.count()

# Held-out ratings of 4 or 5 are treated as relevant for top-10 evaluation.
heldout_relevant = (
    validation.where(F.col("rating") >= 4)
    .join(eval_users.select("user_id"), "user_id")
    .select("user_id", "recipe_id")
    .distinct()
    .cache()
)

# Estimate item popularity from training data for the novelty calculation.
item_popularity = train.groupBy("recipe_id").agg(F.count("*").alias("training_rating_count"))
seen_training = train.select("user_id", "recipe_id").distinct()

# Ask ALS for a larger candidate set, then remove recipes already seen in training.
recommendations = model.recommendForUserSubset(eval_users.select("user_id"), 200)
candidate_recs = (
    recommendations
    .select("user_id", F.explode("recommendations").alias("rec"))
    .select("user_id",
            F.col("rec.recipe_id").cast("int").alias("recipe_id"),
            F.col("rec.rating").cast("double").alias("als_score"))
    .join(seen_training, ["user_id", "recipe_id"], "left_anti")
    .join(item_popularity, "recipe_id", "left")
    .fillna({"training_rating_count": 0})
)
rank_window = Window.partitionBy("user_id").orderBy(F.desc("als_score"))
candidate_recs = (
    candidate_recs
    .withColumn("candidate_rank", F.row_number().over(rank_window))
    .where(F.col("candidate_rank") <= 100)
    .orderBy("user_id", "candidate_rank")
    .cache()
)
candidate_rows = candidate_recs.count()
training_user_count = train.select("user_id").distinct().count()
training_recipe_count = train.select("recipe_id").distinct().count()

# Write compact outputs so the visible R chunks can build the final report tables.
with open(metrics_path, "w", newline="") as f:
    writer = csv.writer(f)
    writer.writerow(["model", "rank", "max_iter", "reg_param", "RMSE", "MAE", "runtime_seconds", "validation_predictions", "status"])
    writer.writerow(["Spark ALS", 20, 10, 0.10, als_RMSE, als_MAE, runtime_seconds, als_prediction_rows, "Completed"])

with open(eval_setup_path, "w", newline="") as f:
    writer = csv.writer(f)
    writer.writerow(["metric", "value"])
    writer.writerow(["training_rows", train_rows])
    writer.writerow(["validation_rows", validation_rows])
    writer.writerow(["evaluation_users", eval_user_count])
    writer.writerow(["candidate_recommendation_rows", candidate_rows])
    writer.writerow(["total_training_ratings", train_rows])
    writer.writerow(["training_user_count", training_user_count])
    writer.writerow(["training_recipe_count", training_recipe_count])

with open(recs_path, "w", newline="") as f:
    writer = csv.writer(f)
    writer.writerow(["user_id", "recipe_id", "als_score", "training_rating_count", "candidate_rank"])
    for row in candidate_recs.collect():
        writer.writerow([row["user_id"], row["recipe_id"], row["als_score"], row["training_rating_count"], row["candidate_rank"]])

with open(heldout_path, "w", newline="") as f:
    writer = csv.writer(f)
    writer.writerow(["user_id", "recipe_id"])
    for row in heldout_relevant.collect():
        writer.writerow([row["user_id"], row["recipe_id"]])

spark.stop()
```

**Full-data Spark ALS Rating Prediction Results**

```{r spark-results, echo=FALSE}
# Read the Spark outputs back into R for reporting.
metrics_path <- file.path(spark_work_dir, "spark_model_metrics.csv")
recs_path <- file.path(spark_work_dir, "spark_candidate_recommendations.csv")
heldout_path <- file.path(spark_work_dir, "spark_heldout_relevant.csv")
eval_setup_path <- file.path(spark_work_dir, "spark_eval_setup.csv")

spark_metrics <- read_csv(metrics_path, show_col_types = FALSE)
spark_eval_setup <- read_csv(eval_setup_path, show_col_types = FALSE)
candidate_recs <- read_csv(recs_path, show_col_types = FALSE)
heldout_relevant <- read_csv(heldout_path, show_col_types = FALSE)

# Compare Spark ALS against a simple bias baseline for rating prediction.
spark_metrics %>%
  transmute(Model = model,
            Rank = if_else(is.na(rank), "NA", format_report_number(rank, 0)),
            `Max iter` = if_else(is.na(max_iter), "NA", format_report_number(max_iter, 0)),
            `Reg param` = if_else(is.na(reg_param), "NA", format_report_number(reg_param, 2)),
            RMSE = format_report_number(RMSE, 3),
            MAE = format_report_number(MAE, 3),
            `Run time seconds` = format_report_number(runtime_seconds, 3),
            `Validation predictions` = format(as.numeric(validation_predictions), big.mark = ","),
            Status = status) %>%
  kable()
```

For the final version, Spark ALS provided the personalized candidate lists that were evaluated with top-N and beyond-accuracy metrics, including coverage, diversity, novelty, and context fit. Therefore, the values were read as the result of the full-data workflow rather than as a separate historical comparison.

**Spark Top-N Evaluation Setup**

```{r eval-setup, echo=FALSE}
# Show the size of the Spark top-N evaluation workflow.
spark_eval_setup %>%
  mutate(metric = dplyr::recode(metric,
                         training_rows = "Training rows",
                         validation_rows = "Validation rows",
                         evaluation_users = "Evaluation users",
                         candidate_recommendation_rows = "Candidate recommendation rows",
                         total_training_ratings = "Total training ratings",
                         training_user_count = "Training user count",
                         training_recipe_count = "Training recipe count"),
         value = format(as.numeric(value), big.mark = ",")) %>%
  transmute(Metric = metric,
            Value = value) %>%
  kable()
```

Compared with Projects 4 and 5, the final project was designed to support the largest practical Food.com interaction set from this sequence. The table below showed the final full-data run alongside the earlier project sizes.

**Project 4, Project 5, and Final Project Modeling Size Comparison**

```{r project-size-comparison, echo=FALSE}
# Compare the earlier modeling sizes with the final raw positive interaction run.
final_training_rows <- as.numeric(spark_eval_setup$value[spark_eval_setup$metric == "training_rows"])
final_validation_rows <- as.numeric(spark_eval_setup$value[spark_eval_setup$metric == "validation_rows"])
final_training_users <- as.numeric(spark_eval_setup$value[spark_eval_setup$metric == "training_user_count"])
final_training_recipes <- as.numeric(spark_eval_setup$value[spark_eval_setup$metric == "training_recipe_count"])

project_size_comparison <- tibble::tribble(
  ~metric, ~project_4, ~project_5, ~current_run,
  "Users", 7795, 24875, final_training_users,
  "Recipes", 8000, 146257, final_training_recipes,
  "Observed/modeling ratings", 253120, 698901, nrow(modeling_interactions),
  "Training ratings", 202496, 559121, final_training_rows,
  "Test or validation ratings", 50624, 123143, final_validation_rows
)

# Print a compact size comparison based on the style used in Project 5.
project_size_comparison %>%
  transmute(Metric = metric,
            `Project 4` = if_else(is.na(project_4), "Not reported", format_report_number(project_4, 0)),
            `Project 5` = if_else(is.na(project_5), "Not reported", format_report_number(project_5, 0)),
            `Final project` = format_report_number(current_run, 0)) %>%
  kable()
```

The utility matrix was still extremely sparse. This was normal for recommender systems because most users only rated a small fraction of all available recipes. Missing entries should be treated as unknown preferences, not dislikes. Otherwise, the model would incorrectly assume that a user rejected thousands of recipes simply because the user never rated them.

**Training Utility Matrix Sparsity**

```{r utility-matrix-sparsity, echo=FALSE}
# Calculate sparsity for the final project's Spark training utility matrix.
possible_pairs <- final_training_users * final_training_recipes
observed_share <- final_training_rows / possible_pairs

# Print the sparsity summary.
tibble(Metric = c("Possible user-recipe pairs",
                  "Possible training matrix entries",
                  "Observed training ratings",
                  "Observed share",
                  "Sparsity"),
       Value = c(format_report_number(possible_pairs, 0),
                 format_report_number(possible_pairs, 0),
                 format_report_number(final_training_rows, 0),
                 format_report_percent(observed_share, 4),
                 format_report_percent(1 - observed_share, 4))) %>%
  kable()
```

# Local Versus Spark Comparison

Project 5 showed that Spark was a scalability tool, not an automatic accuracy improvement. Ultimately, the comparison helped determine whether Spark was needed because the workload had become too large for local tools, not because Spark should be assumed to perform better by default.

**Full-data Same-Split Local and Spark Results**

```{r final-same-split-results, echo=FALSE}
# Combine local model results with Spark ALS results from the same split.
spark_model_results <- spark_metrics %>%
  transmute(Model = model,
            Factors = rank,
            RMSE = RMSE,
            MAE = MAE,
            `Run time seconds` = runtime_seconds,
            `Validation predictions` = as.numeric(validation_predictions),
            Status = status)

same_split_model_results <- bind_rows(local_model_results,
                                      spark_model_results)

# Print the current same-split comparison table.
same_split_model_results %>%
  mutate(Factors = if_else(is.na(Factors), "NA", format_report_number(Factors, 0)),
         RMSE = if_else(is.na(RMSE), "Not completed", format_report_number(RMSE, 3)),
         MAE = if_else(is.na(MAE), "Not completed", format_report_number(MAE, 3)),
         `Run time seconds` = if_else(is.na(`Run time seconds`),
                                      "Not completed",
                                      format_report_number(`Run time seconds`, 3)),
         `Validation predictions` = if_else(is.na(`Validation predictions`),
                                              "Not completed",
                                              format_report_number(`Validation predictions`, 0))) %>%
  kable()
```

The factor count was NA for the bias baseline because that model did not learn latent factors. The run time was reported for the completed models, and any model that did not complete was kept in the table so the limitation was visible. The Spark ALS rerun used the same comparable parameter values as the local factor models: 20 factors, 10 iterations, and 0.10 regularization. Adding local recommenderlab ALS made the comparison more direct because it tested a local ALS-style method rather than only comparing Spark ALS to recosystem. In the separate full-data stress test, the recommenderlab model object was created, and the full rating matrix prediction was attempted. The attempt stopped after about 3 seconds with a LAPACK singular-system error, so RMSE and MAE were not available for that model. The ratings were also highly positive, which helped the simple local bias baseline perform well by learning the overall tendency for users to give high ratings. Therefore, the comparison was fairer after adding the ALS check, but it still showed that Spark was the practical tool for the full-data recommender workflow.

**Local versus Spark Implementation Tradeoffs**

```{r local-spark-comparison, echo=FALSE}
# Compare the role of local tools and Spark in this project sequence.
local_spark_comparison <- tibble::tribble(
  ~dimension, ~local_r_approach, ~spark_als_approach,
  "Setup",
  "Simpler R package workflow used in earlier projects.",
  "Requires Java, PySpark, a Spark session, and Spark DataFrames.",
  "Data scale",
  "Works well for sampled or laptop-scale sparse rating triples.",
  "Better fit when the project uses the largest practical raw interaction file.",
  "Main algorithm role",
  "Good for baselines, fast matrix factorization experiments, and checking whether a local ALS-style model is practical.",
  "Used here as the collaborative-filtering candidate generator.",
  "Strength",
  "Fast iteration and easier debugging.",
  "Scalable joins, distributed ALS, and batch top-N candidate generation.",
  "Weakness",
  "Can run into local memory or runtime limits as data grows.",
  "More operational complexity and not guaranteed to beat simple baselines.",
  "Final project use",
  "Used conceptually as the comparison point from Projects 3, 4, and 5.",
  "Used directly for the final candidate lists that are reranked with context."
)

# Print the local-versus-Spark tradeoff table in the same style as Project 5.
local_spark_comparison %>%
  transmute(Dimension = dimension,
            `Local R approach` = local_r_approach,
            `Spark ALS approach` = spark_als_approach) %>%
  kable()
```

# Context-Aware Reranking

For each synthetic profile, I scored the ALS candidate recipes using available recipe metadata. The final hybrid score combined predicted preference and profile fit. I weighted predicted preference slightly more heavily because the system still needed to recommend recipes the user was likely to enjoy. However, the profile weight was large enough for the context layer to change the list when a recipe was a better fit for the user's nutrition, access, or practicality context. In turn, this made the reranking step a transparent way to examine the tradeoff between preference and context.

```{r reranking}
# Join ALS candidates to recipe item-profile features.
candidates <- candidate_recs %>%
  left_join(recipe_features, by = "recipe_id") %>%
  filter(!is.na(recipe_name)) %>%
  mutate(als_scaled = scale_01(als_score, higher_is_better = TRUE),
         novelty = -log2((training_rating_count + 1) /
                           (as.numeric(spark_eval_setup$value[spark_eval_setup$metric == "total_training_ratings"]) +
                              as.numeric(spark_eval_setup$value[spark_eval_setup$metric == "training_recipe_count"]))))

# Score each candidate recipe against one synthetic profile.
score_profile <- function(candidate_tbl, profile_name) {
  profile_weights <- synthetic_profiles %>%
    filter(profile == profile_name) %>%
    slice(1)

  preference_weight <- profile_weights$preference_weight
  nutrition_weight <- profile_weights$nutrition_weight
  practical_weight <- profile_weights$practical_weight
  restriction_weight <- profile_weights$restriction_weight

  candidate_tbl <- candidate_tbl %>%
    mutate(profile = profile_name,
           nutrition_fit = case_when(
             profile_name == "Heart-conscious" ~
               0.45 * sodium_scaled +
               0.35 * saturated_fat_scaled +
               0.20 * calories_scaled,
             profile_name == "Lower-sugar" ~
               0.55 * sugar_scaled +
               0.20 * as.numeric(!dessert_flag) +
               0.15 * protein_scaled +
               0.10 * calories_scaled,
             profile_name == "High-protein practical meals" ~
               0.55 * protein_scaled +
               0.35 * protein_per_calorie_scaled +
               0.10 * calories_scaled,
             profile_name == "Quick weeknight meals" ~
               0.40 * sodium_scaled +
               0.40 * sugar_scaled +
               0.20 * calories_scaled,
             profile_name == "Budget and access aware" ~
               0.35 * sodium_scaled +
               0.35 * sugar_scaled +
               0.30 * calories_scaled,
             profile_name == "Vegetarian preference" ~
               0.25 * sodium_scaled +
               0.25 * sugar_scaled +
               0.25 * protein_scaled +
               0.25 * calories_scaled,
             TRUE ~ 0.5
           ),
           practical_fit = case_when(
             profile_name == "Quick weeknight meals" ~
               0.45 * minutes_scaled + 0.35 * steps_scaled + 0.20 * ingredients_scaled,
             profile_name == "Budget and access aware" ~
               0.45 * ingredients_scaled + 0.35 * steps_scaled + 0.20 * minutes_scaled,
             profile_name == "High-protein practical meals" ~
               0.45 * ingredients_scaled + 0.35 * minutes_scaled + 0.20 * steps_scaled,
             TRUE ~
               0.40 * ingredients_scaled + 0.30 * minutes_scaled + 0.30 * steps_scaled
           ),
           restriction_fit = case_when(
             profile_name == "Heart-conscious" ~ as.numeric(!processed_meat_flag),
             profile_name == "Lower-sugar" ~ as.numeric(!dessert_flag),
             profile_name == "High-protein practical meals" ~ as.numeric(!dessert_flag),
             profile_name == "Budget and access aware" ~ as.numeric(!specialty_ingredient_flag),
             profile_name == "Vegetarian preference" ~ as.numeric(vegetarian_flag | !meat_flag),
             TRUE ~ 1
           ),
           profile_fit = nutrition_weight * nutrition_fit +
             practical_weight * practical_fit +
             restriction_weight * restriction_fit,
           final_score = preference_weight * als_scaled + profile_fit)
  candidate_tbl
}

# The ALS-only policy keeps the original top-10 ALS ranking.
als_only_lists <- candidates %>%
  group_by(user_id) %>%
  arrange(desc(als_score), .by_group = TRUE) %>%
  slice_head(n = 10) %>%
  ungroup() %>%
  mutate(policy = "ALS only",
         profile = "ALS only",
         final_score = als_scaled,
         profile_fit = NA_real_)

# The hybrid policy reranks the ALS candidates with profile-specific context fit.
hybrid_lists <- map_dfr(synthetic_profiles$profile, ~ score_profile(candidates, .x)) %>%
  group_by(user_id, profile) %>%
  arrange(desc(final_score), desc(als_score), .by_group = TRUE) %>%
  slice_head(n = 10) %>%
  ungroup() %>%
  mutate(policy = "Context-aware hybrid")

recommendation_lists <- bind_rows(als_only_lists, hybrid_lists)
```

# Recommendation-List Evaluation

The top-N evaluation treats held-out recipes rated 4 or 5 as relevant. Since this is an offline evaluation, these metrics are not the same as a real users. However, they are still useful because they show how the list changes when context is added. In particular, they allow me to compare whether the hybrid reranker changes relevance, diversity, novelty, coverage, and practical alignment in ways that appear interpretable.

```{r list-metrics}
# Count each user's held-out relevant recipes.
heldout_counts <- heldout_relevant %>%
  group_by(user_id) %>%
  summarize(relevant_count = n_distinct(recipe_id), .groups = "drop")

# Mark which recommended recipes are hits against the held-out relevant set.
recommended_with_hits <- recommendation_lists %>%
  left_join(heldout_relevant %>% mutate(relevant_hit = TRUE),
            by = c("user_id", "recipe_id")) %>%
  mutate(relevant_hit = replace_na(relevant_hit, FALSE))

# Compute user-level top-10 and beyond-accuracy metrics.
user_list_metrics <- recommended_with_hits %>%
  group_by(policy, profile, user_id) %>%
  summarize(list_n = n(),
            hits = sum(relevant_hit),
            mean_predicted_rating = mean(als_score, na.rm = TRUE),
            mean_profile_fit = mean(profile_fit, na.rm = TRUE),
            mean_sodium = mean(sodium, na.rm = TRUE),
            mean_sugar = mean(sugar, na.rm = TRUE),
            mean_saturated_fat = mean(saturated_fat, na.rm = TRUE),
            mean_protein = mean(protein, na.rm = TRUE),
            mean_minutes = mean(minutes, na.rm = TRUE),
            mean_ingredients = mean(n_ingredients, na.rm = TRUE),
            mean_novelty = mean(novelty, na.rm = TRUE),
            diversity = mean_pairwise_diversity(recipe_id, recipe_features),
            .groups = "drop") %>%
  left_join(heldout_counts, by = "user_id") %>%
  mutate(relevant_count = replace_na(relevant_count, 0),
         precision_at_10 = hits / pmax(list_n, 1),
         recall_at_10 = if_else(relevant_count > 0, hits / relevant_count, NA_real_),
         hit_rate_at_10 = as.numeric(hits > 0))

# Coverage is calculated over the candidate catalog, not the full metadata table.
candidate_pool_size <- n_distinct(candidates$recipe_id)

coverage_summary <- recommendation_lists %>%
  group_by(policy, profile) %>%
  summarize(catalog_coverage = n_distinct(recipe_id) / candidate_pool_size,
            .groups = "drop")

# Average user-level metrics to compare ALS-only and context-aware policies.
list_metric_summary <- user_list_metrics %>%
  group_by(policy, profile) %>%
  summarize(`Precision at 10` = mean(precision_at_10, na.rm = TRUE),
            `Recall at 10` = mean(recall_at_10, na.rm = TRUE),
            `Hit Rate at 10` = mean(hit_rate_at_10, na.rm = TRUE),
            `Mean ALS Score` = mean(mean_predicted_rating, na.rm = TRUE),
            `Mean Diversity` = mean(diversity, na.rm = TRUE),
            `Mean Novelty` = mean(mean_novelty, na.rm = TRUE),
            `Mean Sodium` = mean(mean_sodium, na.rm = TRUE),
            `Mean Sugar` = mean(mean_sugar, na.rm = TRUE),
            `Mean Saturated Fat` = mean(mean_saturated_fat, na.rm = TRUE),
            `Mean Protein` = mean(mean_protein, na.rm = TRUE),
            `Mean Minutes` = mean(mean_minutes, na.rm = TRUE),
            `Mean Ingredients` = mean(mean_ingredients, na.rm = TRUE),
            .groups = "drop") %>%
  left_join(coverage_summary, by = c("policy", "profile"))

```

**Top-N and Beyond-Accuracy Metrics**

```{r list-metrics-table, echo=FALSE}
list_metric_summary %>%
  transmute(Policy = policy,
            Profile = profile,
            `Precision at 10` = format_report_percent(`Precision at 10`, 1),
            `Recall at 10` = format_report_percent(`Recall at 10`, 1),
            `Hit rate at 10` = format_report_percent(`Hit Rate at 10`, 1),
            `Mean ALS score` = format_report_number(`Mean ALS Score`, 3),
            `Mean diversity` = format_report_number(`Mean Diversity`, 3),
            `Mean novelty` = format_report_number(`Mean Novelty`, 3),
            `Mean sodium` = format_report_number(`Mean Sodium`, 3),
            `Mean sugar` = format_report_number(`Mean Sugar`, 3),
            `Mean saturated fat` = format_report_number(`Mean Saturated Fat`, 3),
            `Mean protein` = format_report_number(`Mean Protein`, 3),
            `Mean minutes` = format_report_number(`Mean Minutes`, 3),
            `Mean ingredients` = format_report_number(`Mean Ingredients`, 3),
            `Catalog coverage` = format_report_percent(catalog_coverage, 1)) %>%
  kable()
```

```{r metric-plot, fig.width=9, fig.height=5}
# Plot the beyond-accuracy metrics so policy differences are easier to compare.
list_metric_summary %>%
  select(policy, profile, `Mean Diversity`, `Mean Novelty`, catalog_coverage) %>%
  pivot_longer(cols = c(`Mean Diversity`, `Mean Novelty`, catalog_coverage),
               names_to = "Metric",
               values_to = "Value") %>%
  mutate(Profile_Label = if_else(policy == "ALS only", "ALS only", profile),
         Profile_Label = fct_reorder(Profile_Label, Value, .fun = mean, .desc = FALSE)) %>%
  ggplot(aes(x = Profile_Label, y = Value, fill = Metric)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.65) +
  coord_flip() +
  labs(title = "Beyond-Accuracy Metrics by Recommendation Policy",
       x = NULL,
       y = "Metric Value",
       fill = "Metric") +
  theme_minimal()
```

# Example Recommendations

The tables below show example recommendations for three context-aware profiles. This was a qualitative check. Precision, recall, hit rate, ALS score, diversity, novelty, nutrition, practicality, and coverage were list-level metrics, so I reported them in a small profile-level table before showing the individual recipe examples. The recipe table then focused on the recipe names and features that made the lists easier to interpret.

```{r examples}
# Pick one evaluation user and show a small qualitative recommendation example.
example_user <- recommendation_lists %>%
  count(user_id, sort = TRUE) %>%
  slice(1) %>%
  pull(user_id)

example_profiles <- c("Heart-conscious", "Quick weeknight meals", "High-protein practical meals")
example_metric_profiles <- c("ALS only", example_profiles)
```

**Example Profile-Level Metrics**

```{r example-profile-metrics-table, echo=FALSE}
list_metric_summary %>%
  filter(profile %in% example_metric_profiles) %>%
  mutate(profile = factor(profile, levels = example_metric_profiles)) %>%
  arrange(profile) %>%
  transmute(Profile = profile,
            `Precision at 10` = format_report_percent(`Precision at 10`, 1),
            `Recall at 10` = format_report_percent(`Recall at 10`, 1),
            `Hit rate at 10` = format_report_percent(`Hit Rate at 10`, 1),
            `Mean ALS score` = format_report_number(`Mean ALS Score`, 3),
            `Mean diversity` = format_report_number(`Mean Diversity`, 3),
            `Mean novelty` = format_report_number(`Mean Novelty`, 3),
            `Mean sodium` = format_report_number(`Mean Sodium`, 3),
            `Mean sugar` = format_report_number(`Mean Sugar`, 3),
            `Mean saturated fat` = format_report_number(`Mean Saturated Fat`, 3),
            `Mean protein` = format_report_number(`Mean Protein`, 3),
            `Mean minutes` = format_report_number(`Mean Minutes`, 3),
            `Mean ingredients` = format_report_number(`Mean Ingredients`, 3),
            `Catalog coverage` = format_report_percent(catalog_coverage, 1)) %>%
  kable()
```

**Example Recipe Recommendations**

```{r example-recommendations-table, echo=FALSE}
recommendation_lists %>%
  filter(user_id == example_user,
         profile %in% example_profiles) %>%
  mutate(example_sort_value = case_when(
    profile == "High-protein practical meals" ~ protein,
    profile == "Heart-conscious" ~ saturated_fat,
    profile == "Quick weeknight meals" ~ minutes,
    TRUE ~ final_score
  )) %>%
  group_by(profile) %>%
  arrange(example_sort_value, desc(final_score), .by_group = TRUE) %>%
  slice_head(n = 10) %>%
  ungroup() %>%
  transmute(Profile = profile,
            Recipe = recipe_name,
            `ALS score` = format_report_number(als_score, 3),
            Calories = format_report_number(calories, 1),
            Sodium = format_report_number(sodium, 1),
            Sugar = format_report_number(sugar, 1),
            `Saturated fat` = format_report_number(saturated_fat, 1),
            Protein = format_report_number(protein, 1),
            Minutes = format_report_number(minutes, 0),
            `Number of ingredients` = format_report_number(n_ingredients, 0),
            Novelty = format_report_number(novelty, 2)) %>%
  kable()
```

# Discussion

The final recommender showed why a recipe recommendation system should not be evaluated with one metric. RMSE and MAE were useful because the data included explicit ratings, but they only measured rating prediction. Precision at 10, recall at 10, and hit rate at 10 evaluated the recommendation list against held-out liked recipes. Coverage, diversity, and novelty asked a different question: whether the recommender used the catalog well or fell back to narrow, obvious, repetitive lists. Overall, the results were interpreted as a set of tradeoffs rather than a single winner-take-all score.

For this project, novelty was especially important because a recipe recommender should not only surface the most frequently rated recipes. A popular recipe may be safe from a prediction standpoint, but it may not help a user discover useful new options. Diversity was also important because ten similar recipes could be less useful than a list that gave the user several realistic choices. Coverage mattered because a recommender that only used a small part of the catalog could be technically accurate but limited as a user experience. In a nutrition-focused setting, these beyond-accuracy measures were also connected to autonomy because users may benefit from having multiple feasible options rather than one narrow type of recipe repeated in different forms.

The context-aware hybrid reranker was not meant to replace user preference. Instead, it tried to balance predicted preference with practical health-context fit. In a real application, users should be able to control these goals directly, especially if recommendations are connected to food access, health goals, or dietary restrictions. This project used synthetic profiles because Food.com did not include real health context, allergies, budget, kitchen access, or cultural preference data. As a result, the profiles were understood as a transparent simulation of missing context, not as a substitute for real user-centered design or clinical validation.

# Limitations

- **Synthetic profiles are not real users:** I used guideline-informed, LLM-assisted profiles to make the scenarios realistic, but they are still simulated.
- **Food.com nutrition fields are limited:** Some values are percent daily values rather than full nutrient measurements.
- **Offline metrics are incomplete:** A real recommender would need online testing or user feedback.
- **The holdout is reproducible but not temporal:** Food.com includes dates, and a production-style recommender would ideally train on earlier interactions and test on later interactions. I kept the random 80/20 holdout so the final project stayed comparable to the earlier course projects.
- **Spark ALS still has cold-start limits:** New users and new recipes need content features or onboarding questions.
- **Context weights are transparent but not clinically tuned:** The reranking weights are designed for interpretation, not medical guidance.

# Sources

Food.com Recipes and Interactions. Kaggle dataset by Shuyang Li. <https://www.kaggle.com/datasets/shuyangli94/food-com-recipes-and-user-interactions>

Apache Spark MLlib Collaborative Filtering Documentation. <https://spark.apache.org/docs/latest/ml-collaborative-filtering.html>

U.S. Department of Agriculture and U.S. Department of Health and Human Services. *Dietary Guidelines for Americans, 2020-2025*. 9th Edition. December 2020. <https://www.dietaryguidelines.gov/resources/2020-2025-dietary-guidelines-online-materials>

Koren, Y., Bell, R., & Volinsky, C. (2009). Matrix factorization techniques for recommender systems. *Computer*, 42(8), 30-37. <https://datajobs.com/data-science-repo/Recommender-Systems-%5BNetflix%5D.pdf>

Trattner, C., & Elsweiler, D. (2017). *Food recommender systems: Important contributions, challenges and future research directions*. <https://arxiv.org/abs/1711.02760>

Figueroa, C. A., Torkamaan, H., Bhattacharjee, A., Hauptmann, H., Guan, K. W., & Sedrakyan, G. (2025). Designing health recommender systems to promote health equity: A socioecological perspective. *Journal of Medical Internet Research*, 27, e60138. <https://doi.org/10.2196/60138>
